---
title: "Week 3 Exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 3.4(a)

We need to compute the integral \mathbb{E}_f(h(X)), where $X \sim \mathcal{N}(0, 1)$ and
$$
h(x) = \exp\left( 
  -\frac{(x-3)^2}{2}
\right)
+
\exp\left( 
  -\frac{(x-6)^2}{2}
\right).
$$
Remember that the PDF of a standard normal is given by
$$
f(x) = \frac{1}{\sqrt{2\pi}}\exp\left( 
  -\frac{x^2}{2}
\right).
$$
It will be helpful to abbreviate these kinds of integrals.
Introduce the notation
$$
q(x; \mu, \sigma^2) = \exp\left( 
  -\frac{(x-\mu)^2}{2\sigma^2}
\right).
$$
It will be helpful later to notice that this is proportional to the PDF of a Gaussian
random variable with mean $\mu$ and variance $\sigma^2$.
So $f(x) = \frac{1}{\sqrt{2 \pi}} N(x; 0, 1)$ and $h(x) = N(x; 3, 1) + N(x; 6, 1)$.

Now we have that
$$\begin{aligned}
\mathbb{E}_f(h(X)) &= \int_{-\infty}^\infty h(x) f(x) dx \\
&= \frac{1}{\sqrt{2 \pi}}\int_{-\infty}^\infty [N(x; 3, 1) + N(x; 6, 1)] N(x; 0, 1) dx .
\end{aligned}$$
It looks like rearranging terms of the form $N(x; \mu, 1) N(x; 0,1)$ will be helpful, so
let's take a closer look at those.

$$\begin{aligned}
q(x; \mu, 1) q(x; 0, 1) &= \exp\left(
  -\frac{(x - \mu)^2}{2}
\right)
\exp\left(
  -\frac{x^2}{2}
\right) \\
&= \exp\left(
  -\frac{1}{2}(x^2 - 2\mu x + \mu^2 + x^2)
\right) \\
&= \exp\left(
  -\frac{1}{2}(2x^2 - 2\mu x + \mu^2)
\right) \\
&= \exp\left(
  -x^2 - \mu x + \frac{1}{2}\mu^2
\right) \\
&= \exp\left(
  -\left(x^2 - \frac{1}{2} \mu\right)^2 + \frac{\mu^2}{4}
\right) \\
&= \exp\left(
  -\left(x^2 - \frac{1}{2} \mu\right)^2
\right)
\exp\left(
-\frac{\mu^2}{4}
\right)\\
&= q\left(x; \frac{1}{2}\mu, \frac{1}{2}\right) \exp\left( -\frac{\mu^2}{4}\right)
\end{aligned}$$

Now let's think about integrating this
$$\begin{aligned}
\frac{1}{\sqrt{2\pi}}\int_{-\infty}^\infty q(x; \mu, 1) q(x; 0, 1) dx 
&= \frac{1}{\sqrt{2\pi}}\exp\left(-\frac{\mu^2}{4}\right) 
  \int_{-\infty}^\infty q\left(x; \frac{1}{2}\mu, \frac{1}{2}\right) dx
\end{aligned}$$
The integral on the right above is easy, because we know $q(x; \frac{1}{2}\mu, 2)$ is
proportional to the PDF of a Gaussian with mean $\frac{1}{2}\mu$ and variance $2$.
Moreover, we know that such PDFs integrate to 1, precisely because they are PDFs.
If we look at the Gaussian PDF, we see that
$$\begin{aligned}
  N(x; \mu, \sigma^2) &= \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(
    -\frac{(x-\mu)^2}{2\sigma^2}
  \right) \\
  &= \frac{1}{\sqrt{2\pi\sigma^2}}q(x; \mu, \sigma^2)
\end{aligned}$$
Since the left-hand-side integrates to 1, we must have that
$$
\int_{-\infty}^\infty q(x; \mu, \sigma^2) dx = \sqrt{2\pi\sigma^2}.
$$
Therefore
$$\begin{aligned}
\frac{1}{\sqrt{2\pi}}\int_{-\infty}^\infty q(x; \mu, 1) q(x; 0, 1) dx
&= \frac{\sqrt{\pi}}{\sqrt{2\pi}}\exp\left(-\frac{\mu^2}{4}\right) \\
&=\frac{1}{\sqrt{2}}\exp\left(-\frac{\mu^2}{4}\right)
\end{aligned}$$
So that
$$
E_f[h(X)] = \frac{1}{\sqrt{2}}\left[
  \exp\left(-\frac{9}{4}\right) + \exp\left(-\frac{36}{4}\right)
\right]
$$

# Exercise 3.4(b)
```{r sampling_graph}
  library(ggplot2)
  set.seed(123)
  n = 10000
  h <- function(x) {
    exp(-0.5*(x-3)^2) + exp(-0.5*(x-6)^2)
  }
  truth = 1/sqrt(2)*(exp(-9/4) + exp(-36/4))
  
  x_sample = rnorm(n)
  h_sample = h(x_sample)
  h_bar_n = 1/1:n * cumsum(h_sample)
  df = data.frame(n=1:n, h_bar_n=h_bar_n)
  ggplot(data=df, aes(x=n, y=h_bar_n, group=1)) +
    geom_line(color="red") +
    geom_hline(yintercept=truth, linetype="dashed", color="blue") +
    ylab(expression(bar(h)[n]))
```

# Exercise 3.4(c)
```{r sampling_graph_nonconvergent}
  library(ggplot2)
  set.seed(123)
  n = 10000
  h <- function(x) {
    exp(-0.5*(x-3)^2) + exp(-0.5*(x-6)^2)
  }
  truth = 1/sqrt(2)*(exp(-9/4) + exp(-36/4))
  
  x_sample = runif(n, -8,-1)
  f <- function(x) {
    1/sqrt(2*pi) * exp(-0.5*x^2)
  }
  g <- function(x) {
    1/7
  }
  weights = f(x_sample) / g(x_sample)
  h_sample = h(x_sample)
  h_bar_n = 1/1:n * cumsum(h_sample*weights)
  df = data.frame(n=1:n, h_bar_n=h_bar_n)
  ggplot(data=df, aes(x=n, y=h_bar_n, group=1)) +
    geom_line(color="red") +
    geom_hline(yintercept=truth, linetype="dashed", color="blue") +
    ylab(expression(bar(h)[n]))
```

# Exercise 3.5
```{r sampling_graph_3.5}
  library(ggplot2)
  set.seed(123)
  n = 10000
  lambda= 1
  h <- function(x) {
    exp(-0.5*(x-3)^2) + exp(-0.5*(x-6)^2)
  }
  truth = 1/sqrt(2)*(exp(-9/4) + exp(-36/4))
  
  x_sample = rexp(n, lambda)
  f <- function(x) {
    1/sqrt(2*pi) * exp(-0.5*x^2)
  }
  g <- function(x) {
    lambda*exp(-lambda*x)
  }
  weights = f(x_sample) / g(x_sample)
  h_sample = h(x_sample)
  h_bar_n = 1/1:n * cumsum(h_sample*weights)
  df = data.frame(n=1:n, h_bar_n=h_bar_n)
  ggplot(data=df, aes(x=n, y=h_bar_n, group=1)) +
    geom_line(color="red") +
    geom_hline(yintercept=truth, linetype="dashed", color="blue") +
    ylab(expression(bar(h)[n]))
```